 

### 2021-07-10 vertical-thermal-review.Rmd
edits to Tleaf section
@NidhiVinod , we need to have a discussion to reconcile our two versions, #48

Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by the other biophysical drivers reviewed above and by leaf traits (see next section), with leaves adjusted to their environment so as to maintain, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]. 
Leaves can often be either cooler than air at high $T_{air}$ or warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ (megathermy) [@doughtyAreTropicalForests2008].
However, rarely is $T_{leaf}$ equal to $T_{air}$ [poikilothermy, @cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].

@@

**attempted to oraganize**

~~Solar and longwave radiation can dramatically elevate $T_{leaf}$ well above $T_{air}$ (Fig. `r fig_leaf_T`a) because the heat absorbed through high irradiance can be greater than heat lost through sensible heat (via $ga$ ie. boundary layer conductance) and latent heat loss (via $gs$ ie. stomatal conductance and $E$ ie evaporation)[@bonan_ecological_2016].

$Ga$ and $Gs$ are dominant factors in $T_{leaf}$ flux. $Ga$ increases with wind speed which also in turn increases $gs$, thereby reduces $T_{leaf}$-$T_{air}$ (Fig. `r fig_leaf_T`b). Therefore, high windspeeds in upper canopies such as in NEON sites with windspeeds ~4 ms (Fig. `r fig_NEON_vertical`) are capable of lowering $T_{leaf}-T_{air}$ that are exposed to high radiation loads, upto ~ 4$^\circ$C (Fig. `r fig_leaf_T`b)[@bonan_ecological_2016; @PhotosyntheticAcclimationSimultaneous2004]. Low wind speeds typically decouple leaves from $T_{air}$, more so when solar radiation is high and $gs$ is low. Additionally, $T_{leaf}-T_{air}$ increases with humidity (Fig. `r fig_leaf_T`c) because reduced $ga$ stifles $E$ and $gs$ that dissapate heat [@songTreeSurfaceTemperature2020].

Latent heat loss through maximum $gs$, cools the leaf and decreases $T_{leaf}$-$T_{air}$, allowing coupling between leaf and air. However in conditions of low soil moisture during drought (Fig. `r fig_leaf_T`), greater temperature than $gs$ thermal optimum, and air dryness causing an increase in VPD, $gs$ becomes limited, thereby increasing $T_{leaf}$-$T_{air}$ [@bonan_ecological_2016]. Limitation to $gs$ can also exaggerate effects of biophysical factors that contribute to energy inputs such as, solar radiation, humidity and $T_{air}$ (Fig. `r fig_leaf_T`)[@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @dongBiophysicalHomoeostasisLeaf2017].

Leaf size and physiology [see next section] influence $ga$, $gs$, leaf thermal capacitance (capacity to hold heat), and therefore also $T_{leaf}$-$T_{air}$. Smaller leaves have thinner boundary layer and higher $ga$, that enables coupling with $T_{air}$ compared to large leaves that can be more decoupled from $T_{air}$ (Fig. `r fig_leaf_T` a). Therefore, smaller leaves with adequate water, will be most effective at cooling during high temperatures than larger leaves [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007;@fauset_differences_2018]. Additionally, under low raditation, large leaves can be cooler than small leaves due to greater $gs$ [(Fig. `r fig_leaf_T` d; @leighThickLeavesAvoid2012;@bonan_ecological_2016]. Sun leaves, such as in the upper canopies, have greater $gs$, due to greater stomatal density [Table 1] and lower thermal capacitance due to lower water potential than shade leaves. However, exposure to high irradiance in upper canopies which is a dominant source of energy input increases $T_{leaf}$-$T_{leaf}$ more than shade leaves that are shielded from biophysical conditions [@millerOnlySunlitLeaves2021]. However, because shade leaves have higher thermal capacitance and comparitevly lower $gs$, when exposed to an increase in radiation or $T_{air}$, $T_{leaf}$-$T_{air}$ can easily become more elevated than sun leaves (Fig. `r fig_leaf_T`a)[@schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].

@@

@@

### In situ observations
**Following the principles outlined above, vertical microclimate profiles, leaf traits (see "Leaf Traits" section), and $g_s$ (see "Leaf metabolism and thermal responses" section) together explain in-situ $T_{leaf}$ observations (Figs. `r fig_NEON_vertical`-`r fig_leaf_T`).** 
Often, $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s$* / ~due to high evapotranspiration~, and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).
In closed-canopy forests, including tropical and temperate broadleaf forests, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optima for photosynthesis ($T_{opt}$) and $T_{air}$, because of $g_s$-limitation and high irradiance [@niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018]. 
In contrast, in open forests with lower $LAI$, lower-canopy $T_{leaf}$ can be greater than upper canopy $T_{leaf}$, due to greater light permeability, *vertical air mixing*, and still air in the lower canopy and heating from the ground [Figs. `r fig_NEON_vertical`h, `r SIfig_NEON_all`b,e, @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016]. 
Similarly, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019]. 
Greater $RH$ in closed-inner canopy can also increase $T_{leaf}$ relatively more than upper-canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].

Spatial distribution of canopy temperature also shows that exposed tropical tree barks can be much warmer than leaves and flowers [@pauTropicalForestTemperature2018].

>>>>

Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by other biophysical drivers (reviewed above) and by leaf traits and stomatal conductance (see following sections), with leaves adjusted to their environment so as to maintain, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]. 
Leaves can be either cooler than air at high $T_{air}$ or warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ [megathermy, @doughtyAreTropicalForests2008].
While $T_{leaf}$ is most commonly within a few degrees of $T_{air}$ [poikilothermy, @cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020], it is rarely exactly equal.

@@

**Nidhi's rearrangement:**

~~Solar and longwave radiation can dramatically elevate $T_{leaf}$ well above $T_{air}$ (Fig. `r fig_leaf_T`a) because the heat absorbed through high irradiance can be greater than heat lost through sensible heat (via $g_a$; *i.e.*, boundary layer conductance) and $\lambda E$ (via $g_s$ ie. stomatal conductance and $E$ ie evaporation)[@bonan_ecological_2016].

$g_a$ and $g_s$ are dominant factors in $T_{leaf}$ flux. $Ga$ increases with wind speed which also in turn increases $gs$, thereby reduces $T_{leaf}$-$T_{air}$ (Fig. `r fig_leaf_T`b). Therefore, high windspeeds in upper canopies such as in NEON sites with windspeeds ~4 ms (Fig. `r fig_NEON_vertical`) are capable of lowering $T_{leaf}-T_{air}$ that are exposed to high radiation loads, upto ~ 4$^\circ$C (Fig. `r fig_leaf_T`b)[@bonan_ecological_2016; @PhotosyntheticAcclimationSimultaneous2004]. Low wind speeds typically decouple leaves from $T_{air}$, more so when solar radiation is high and $gs$ is low. Additionally, $T_{leaf}-T_{air}$ increases with humidity (Fig. `r fig_leaf_T`c) because reduced $ga$ stifles $E$ and $gs$ that dissapate heat [@songTreeSurfaceTemperature2020].

Latent heat loss through maximum $g_s$, cools the leaf and decreases $T_{leaf}$-$T_{air}$, allowing coupling between leaf and air. However in conditions of low soil moisture during drought (Fig. `r fig_leaf_T`), greater temperature than $gs$ thermal optimum, and air dryness causing an increase in VPD, $gs$ becomes limited, thereby increasing $T_{leaf}$-$T_{air}$ [@bonan_ecological_2016]. Limitation to $gs$ can also exaggerate effects of biophysical factors that contribute to energy inputs such as, solar radiation, humidity and $T_{air}$ (Fig. `r fig_leaf_T`)[@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @dongBiophysicalHomoeostasisLeaf2017].

Leaf size and physiology [see next section] influence $g_a$, $g_s$, leaf thermal capacitance (capacity to hold heat), and therefore also $T_{leaf}$-$T_{air}$. Smaller leaves have thinner boundary layer and higher $g_a$, that enables coupling with $T_{air}$ compared to large leaves that can be more decoupled from $T_{air}$ (Fig. `r fig_leaf_T` a). Therefore, smaller leaves with adequate water, will be most effective at cooling during high temperatures than larger leaves [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007;@fauset_differences_2018]. Additionally, under low raditation, large leaves can be cooler than small leaves due to greater $gs$ [(Fig. `r fig_leaf_T` d; @leighThickLeavesAvoid2012;@bonan_ecological_2016]. Sun leaves, such as in the upper canopies, have greater $gs$, due to greater stomatal density [Table 1] and lower thermal capacitance due to lower water potential than shade leaves. However, exposure to high irradiance in upper canopies which is a dominant source of energy input increases $T_{leaf}$-$T_{leaf}$ more than shade leaves that are shielded from biophysical conditions [@millerOnlySunlitLeaves2021]. However, because shade leaves have higher thermal capacitance and comparitevly lower $gs$, when exposed to an increase in radiation or $T_{air}$, $T_{leaf}$-$T_{air}$ can easily become more elevated than sun leaves (Fig. `r fig_leaf_T`a)[@schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].

@@

**Krista's version:**

@@

### In situ observations
**Following the principles outlined above, vertical microclimate profiles, leaf traits (see "Leaf Traits" section), and $g_s$ (see "Leaf metabolism and thermal responses" section) together explain in-situ $T_{leaf}$ observations.** 
Often, $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s$* / ~due to high evapotranspiration~, and therefore differs little between understory and canopy.
In forests with dense canopies, including tropical and temperate broadleaf forests, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optima for photosynthesis ($T_{opt}$) and $T_{air}$, because of $g_s$-limitation and high irradiance [@niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018]. 
In contrast, in open forests with lower $LAI$, lower-canopy $T_{leaf}$ can equal or even exceed upper canopy $T_{leaf}$, due to greater light permeability, vertical air mixing, and still air in the lower canopy and heating from the ground [Figs. `r fig_NEON_vertical`h, `r SIfig_NEON_all`b,e, @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016]. 
Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019]. 
Greater $RH$ in closed-inner canopy can also increase $T_{leaf}$ relatively more than upper-canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].

It is noteworthy that $T_{leaf}$ often differs substantially from that of branches and stems. 
For instance, exposed tropical tree bark can be much warmer than leaves and flowers [@pauTropicalForestTemperature2018]. 
Thus, the vegetation temperature ($T_{veg}$) measured using sensors that integrate across both leaves and woody vegetation (e.g., infrared sensors for $T_{veg}$ at NEON sites, Fig. `r fig_NEON_vertical`h) does not always equal $T_{leaf}$, particularly in settings where leaf area is low relative to woody vegetation (*i.e.*, low-LAI ecosystems and understories). 
Recent advances in $T_{leaf}$ modeling (see "Scaling across Space and Time") will enhance our understanding of large scale $T_{leaf}$ vertical patterns [@stillImagingCanopyTemperature].
